Combinatorics of the geometry of Wilson loop diagrams II: Grassmann necklaces, dimensions, and denominators

### 2019-04-09 GWLDCII.tex
paper 2 positroid background

@@

\begin{dfn} \label{VPropdfn}
Let $W = (\cP, V)$ be a Wilson loop diagram.
\begin{enumerate}
\item For $p = (i,j) \in \cP$, let $V(p) = \{i, i+1, j, j+1\}$ denote the set of vertices supporting $p$. For a set of propagators $P \subseteq \cP$, define 
\[V(P) = \bigcup_{p \in P} V(p)\]
to be the vertex support of $P$.
\item For $U \subseteq V$, the set of propagators supported on $U$ is denoted by
\[\Prop(U) = \{ p \in \cP\ |\ V(p) \cap U \neq \emptyset \}.\]
\item Vertices of $W$ that do not support any propagators are called non-supporting. The set of all non-supporting vertices in $W$ is denoted $F(\emptyset)$. 
\end{enumerate}
\end{dfn}

\begin{dfn} \label{subdiagramdfn}
Let $W = (\cP, V)$ be an admissible Wilson loop diagram. Then $W' = (Q,U)$ is a subdiagram of $W$ if 
\[ Q \subseteq \cP \quad \text{ and }\quad V(Q) \subseteq U \subseteq V\;.\]
\end{dfn}

@@

The first condition bounds the total number of propagators in the diagram, while the second limits how densely propagators can be fitted into any part of the diagram; in particular, it prohibits both propagators that connect two adjacent edges and parallel propagators that start and end on the same pair of edges. The third condition simply requires that propagators are non-crossing.

These conditions are imposed by the physical interpretation of Wilson loop diagrams (see for example [ref]). We therefore restrict our attention in this paper to admissible Wilson loop diagrams and their subdiagrams only.

@@

\begin{eg}
Consider the admissible Wilson loop diagram 
\[W = (\{(1,4), (2,4), (5,7), (5,8)\},[8]),\]
which is represented graphically by the diagram
\[W = \begin{tikzpicture}[rotate=67.5,baseline=(current bounding box.east)]
	\begin{scope}
	\drawWLD{8}{1.5}
	\drawnumbers
	\drawprop{1}{0}{4}{0}
	\drawprop{2}{0}{4}{-1}
    \drawprop{5}{0}{8}{0}
    \drawprop{5}{1}{7}{0}
		\end{scope}
	\end{tikzpicture}\]
Ordering the propagators as listed above, we obtain the associated matrix
\[ C(W) = \left(
\begin{array}{cccccccc}
c_{1,1} & c_{1,2} & 0 & c_{1,4} & c_{1,5} & 0 & 0 & 0 \\
0 & c_{2,2} & c_{2,3} & c_{2,4} & c_{2,5} & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & c_{3,5} & c_{3,6} & c_{3,7} & c_{3,8} \\
c_{4,1} & 0 & 0 & 0 & c_{4,5} & c_{4,6} & 0 & c_{4,8}  \\
\end{array}
\right) \;.\]
\end{eg}

The matrix $C(W)$ parametrizes a subspace of $\Gr(k, n)$ as shown in \cite{wilsonloop}; we denote this subspace by $\Sigma(W)$.

From $C(W)$ we obtain the matroid $M(W)$ associated to $W$, which is simply the matroid represented by $C(W)$: its ground set is the set $V$ of vertices of $W$, and the independent sets are exactly the sets $U \subseteq V$ such that the columns of $C(W)$ indexed by $U$ are linearly independent. More detail on this is given in Section~\ref{sec:positroid background} below.

The Wilson loop diagram also define a volume form on $\Sigma(W)$, namely 
\[ \Omega(W) = \frac{\prod_{r=1}^{|\cP|} \prod_{v \in V_{p_r}} \textrm{d}c_{p_r}}{R(W)} \;. \]
The interested reader is referred to [ref] for more information about the volume form. Of interest to us in this paper is the denominator $R(W)$, which we define next.

\begin{dfn}\label{def R(W)}
Let $W = (\cP,V)$ be an admissible Wilson loop diagram, and fix an edge $e \in V$. Let $\{p_1 \ldots p_r\}$ be the propagators with one end lying on edge $e$, ordered so that $p_1$ is closest to the vertex $e$, $p_r$ closest to $e+1$, and $p_i$ is closer to $e$ than $p_{i+1}$ is. Then we define 
\[ R_e =  c_{p_1,e+1} \prod_{j= 1}^{r-1} \left((c_{p_j,e} c_{p_{j+1},e+1} - c_{p_{j+1},e} c_{p_{j},e+1} ) \right) c_{p_r,e}\;.\]
Note that if $r = 1$ (i.e. there is exactly one propagator lying on edge $e$) then $R_e = c_{p,e} c_{p,e+1}$. The denominator $R(W)$ is defined to be the product of all of the $R_e$, i.e. 
\[R(W) = \prod_{e \in V} R_e.\]
\end{dfn}
It is this polynomial $R(W)$ that we characterize in Section~\ref{sec poles}, where we give a simple algorithm to read $R(W)$ directly from the Wilson loop diagram.

\subsection{Positroid background}\label{sec:positroid background}

[positroids]

Let $\binom{[n]}{k}$ be the set of all $k$-subsets of the cyclically ordered set $[n]$.  For each $j \in [n]$, we can define a total order $\leq_j$ on the interval $[n]$ by
\[ j <_j j+1 <_j \dots <_j n <_j 1 \dots <_j j-1\;.\]
This in turn induces a total order on $\binom{[n]}{k}$, namely the lexicographic order with respect to $<_j$.  It also induces a separate partial order $\gale{j}$ on $\binom{[n]}{k}$ (the {\bf Gale order} \cite{Gale}), which is defined as follows: for 
\[A = \{a_1 <_j a_2 <_j \dots <_j a_k\} \text{ and } B = \{b_1 <_j b_2 <_j \dots <_j b_k\} \in \binom{[n]}{k},\] we define
\[A \gale{j} B \text{ if and only if } a_r \leq_j b_r \text{ for all }1 \leq r \leq k.\]
For example, in $\binom{[6]}{3}$ we have $\{2,5,6\}\gale{2} \{2,6,1\}$ but $\{2,5,6\}\not\gale{2}\{3,4,6\}$.

@@

By \cite[Theorem 17.1]{Postnikov}, the Grassmann necklaces of type $(k,n)$ are in 1-1 correspondence with the positroid cells in $\Gr(k,n)$.  This correspondence is given explicitly in \cite[Theorem 8]{Oh}: if $(I_1, \dots, I_n)$ is the Grassmann necklace associated to a positroid $M = ([n],\cB)$, then the bases of $M$ are exactly
\[\cB = \left\{J \in \binom{[n]}{k}\ :\ I_i \gale{i} J \ \forall i \in [n]\right\}.\]

@@

By \cite[Theorem 6.5]{Postnikov}, the set of all Le diagrams that fit within a $k\times(n-k)$ rectangle is in 1-1 correspondence with the positroid cells of $\Gr(k,n)$. The dimension of a positroid cell is equal to the number of $+$ squares in its Le diagram \cite[Theorem 6.5]{Postnikov}.

@@

@@

In this section we will give an algorithm to go directly from a Wilson loop diagram to the Grassmann necklace of its positroid.  As well as the result itself, the techniques developped will be useful throughout.  One of the key insights is that an element of the Grassmann necklace can be viewed as a function from the propagators of the diagram to the vertices in the Grassmann necklace element.  This is captured in Definition~\ref{def I_i as a function}, and used throughout the paper.

>>>>

\begin{figure}[h]
\[W\ =\ \begin{tikzpicture}[rotate=67.5,baseline=(current bounding box.east)]
	\begin{scope}
	\drawWLD{8}{1.5}
	\drawnumbers
	\drawprop{1}{0}{4}{0}
	\drawprop{2}{0}{4}{-1}
    \drawprop{5}{0}{8}{0}
    \drawprop{5}{1}{7}{0}
		\end{scope}
	\end{tikzpicture}\]

\caption{The Wilson loop diagram $W = \big(\{(1,4), (2,4), (5,7), (5,8)\},[8]\big)$ which has four propagators and eight vertices.}
\label{fig:ex of WLD}
\end{figure}

@@

\begin{dfn} \label{VPropdfn}
Let $W = (\cP, V)$ be a Wilson loop diagram.
\begin{enumerate}
\item For $p = (i,j) \in \cP$, let $V(p) = \{i, i+1, j, j+1\}$ denote the set of vertices supporting $p$. For a set of propagators $P \subseteq \cP$, define 
\[V(P) = \bigcup_{p \in P} V(p)\]
to be the vertex support of $P$.
\item For $U \subseteq V$, the set of propagators supported on $U$ is denoted by
\[\Prop(U) = \{ p \in \cP\ |\ V(p) \cap U \neq \emptyset \}.\]
Note that a propagator does not need to have its entire support contained in $U$. $\Prop(U)$ is the set of all propagators that would be affected by removing the set $U$.
\item Vertices of $W$ that do not support any propagators are called {\bf non-supporting}. The set of all non-supporting vertices in $W$ is denoted $F(\emptyset)$. 
\end{enumerate}
\end{dfn}

\begin{dfn} \label{subdiagramdfn}
Let $W = (\cP, V)$ be an admissible Wilson loop diagram. Then $W' = (Q,U)$ is a {\bf subdiagram} of $W$ if $Q \subseteq \cP$ and $V(Q) \subseteq U \subseteq V$.
\end{dfn}

@@

The first condition bounds the total number of propagators in the diagram, while the second limits how densely propagators can be fitted into any part of the diagram; in particular, it prohibits propagators that connect two adjacent edges and parallel propagators that start and end on the same pair of edges. The third condition simply requires that propagators are non-crossing.

These conditions are imposed by the physical interpretation of Wilson loop diagrams (see for example \hlfix{[ref]}{probably for Susama: is there a sensible paper to cite here?}). We therefore restrict our attention in this paper to admissible Wilson loop diagrams and their subdiagrams only.

@@

\begin{eg}
Consider the admissible Wilson loop diagram 
\[W = \big(\{(1,4), (2,4), (5,7), (5,8)\},[8]\big),\]
from Figure~\ref{fig:ex of WLD}. Ordering the propagators as listed above, we obtain the associated matrix
\[ C(W) = \left(
\begin{array}{cccccccc}
c_{1,1} & c_{1,2} & 0 & c_{1,4} & c_{1,5} & 0 & 0 & 0 \\
0 & c_{2,2} & c_{2,3} & c_{2,4} & c_{2,5} & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & c_{3,5} & c_{3,6} & c_{3,7} & c_{3,8} \\
c_{4,1} & 0 & 0 & 0 & c_{4,5} & c_{4,6} & 0 & c_{4,8}  \\
\end{array}
\right) \;.\]
\end{eg}

The matrix $C(W)$ parametrizes a subspace of $\Gr(k, n)$, as shown in \cite{wilsonloop}; we denote this subspace by $\Sigma(W)$.

From $C(W)$ we obtain the matroid $M(W)$ associated to $W$, which is simply the matroid represented by $C(W)$: its ground set is the set $V$ of vertices of $W$, and the independent sets are exactly the sets $U \subseteq V$ such that the columns of $C(W)$ indexed by $U$ are linearly independent. More detail on this is given in Section~\ref{sec:M(W)} below.

The Wilson loop diagram also defines a volume form on $\Sigma(W)$, namely 
\[ \Omega(W) = \frac{\prod_{r=1}^{|\cP|} \prod_{v \in V_{p_r}} \textrm{d}c_{p_r}}{R(W)} \;. \]
The interested reader is referred to \hlfix{[ref]}{any paper that talks about the volume form} for more information about the volume form. Of interest to us in this paper is the denominator $R(W)$, which we define next.

\begin{dfn}\label{def R(W)}
Let $W = (\cP,V)$ be an admissible Wilson loop diagram, and fix an edge $e$. Let $\{p_1 \ldots p_r\}$ be the propagators with one end lying on edge $e$, ordered so that $p_1$ is closest to the vertex $e$, $p_r$ closest to $e+1$, and $p_i$ is closer to $e$ than $p_{i+1}$ is. Define 
\[ R_e =  c_{p_1,e+1} \prod_{j= 1}^{r-1} \left((c_{p_j,e} c_{p_{j+1},e+1} - c_{p_{j+1},e} c_{p_{j},e+1} ) \right) c_{p_r,e}\;\]
to be the component of $R(W)$ associated to edge $e$. Note that if $r = 1$ (i.e. there is exactly one propagator lying on edge $e$) then this expression simplifies to $R_e = c_{p,e} c_{p,e+1}$. The denominator $R(W)$ is defined to be the product of all of the $R_e$, i.e. 
\[R(W) = \prod_{e \in V} R_e.\]
\end{dfn}
Understanding the polynomial $R(W)$ (and hence the singularities of the volume form $\Omega(W)$) \hlfix{is important for the physics}{help?}. In Section~\ref{sec poles} we give a simple algorithm to read $R(W)$ directly from the Wilson loop diagram.

\subsection{Positroids, Grassmann necklaces, and Le diagrams}\label{sec:positroid background}

Let $\binom{[n]}{k}$ denote the set of all $k$-subsets of the cyclically ordered set $[n]$.

For our purposes, a {\bf positroid} is a matroid $M$ with ground set $[n]$ and bases $\cB \subseteq \binom{[n]}{k}$ which can be represented by an element of the positive Grassmannian $\Gr(k,n)$, i.e. such that there exists a real full-rank $k\times n$ matrix whose $k\times k$ minors are all nonnegative and such that the minor $[[k]|J]$ is positive if and only if $J \in \cB$.

The interested reader is referred to \cite{Postnikov} for an introduction to the geometry and combinatorics of positroid cells in the positive Grassmannian. Postnikov shows in \cite{Postnikov} that the positroids of $\Gr(k,n)$ are indexed by many interesting collections of objects, including Le diagrams, Grassmann necklaces, decorated permutations, and plabic graphs. The two most suited to our purposes are the Grassmann necklaces and the Le diagrams, which we introduce below. In order to do so, we first need some preliminary definitions.

For each $j \in [n]$, we can define a total order $\leq_j$ on the interval $[n]$ by
\[ j <_j j+1 <_j \dots <_j n <_j 1 \dots <_j j-1\;.\]
This in turn induces a total order on $\binom{[n]}{k}$, namely the lexicographic order with respect to $<_j$.  It also induces a separate partial order $\gale{j}$ on $\binom{[n]}{k}$ (the {\bf Gale order} \cite{Gale}), which is defined as follows: for 
\[A = \{a_1 <_j a_2 <_j \dots <_j a_k\} \text{ and } B = \{b_1 <_j b_2 <_j \dots <_j b_k\} \in \binom{[n]}{k},\] we define
\[A \gale{j} B \text{ if and only if } a_r \leq_j b_r \text{ for all }1 \leq r \leq k.\]
For example, in $\binom{[6]}{3}$ we have $\{2,5,6\}\gale{2} \{2,6,1\}$ but $\{2,5,6\}\not\gale{2}\{3,4,6\}$.

@@

By \cite[Theorem 17.1]{Postnikov}, the Grassmann necklaces of type $(k,n)$ are in 1-1 correspondence with the positroid cells in $\Gr(k,n)$.  This correspondence is given explicitly in \cite[Theorem 8]{Oh}: if $(I_1, \dots, I_n)$ is the Grassmann necklace associated to a positroid $M = ([n],\cB)$, then the bases of $M$ are exactly
\[\cB = \left\{J \in \binom{[n]}{k}\ :\ I_i \gale{i} J \ \forall i \in [n]\right\}.\]
Thus the Grassmann necklace is well suited to testing whether a given $k$-set is a basis for $M$, or for generating a list of all bases of $M$.

@@

By \cite[Theorem 6.5]{Postnikov}, the set of all Le diagrams that fit within a $k\times(n-k)$ rectangle is in 1-1 correspondence with the positroid cells of $\Gr(k,n)$. Le diagrams are particularly useful for comparing dimensions of positroids, since the dimension of a positroid is equal to the number of $+$ squares in its Le diagram \cite[Theorem 6.5]{Postnikov}.

@@

\subsection{The positroid associated to an admissible Wilson loop diagram}\label{sec:M(W)}

words go here

@@

In this section we will give an algorithm to go directly from a Wilson loop diagram to the Grassmann necklace of its positroid. This is a useful result in and of itself, as it greatly simplifies the process of identifying the positroid associated to a given WLD. It also allows us to examine how the positroid changes as we make minor changes to the Wilson loop diagram; indeed, the techniques we develop in this section to verify that our algorithm produces the correct Grassmann necklace will also be used repeatedly in later sections.

One of the key insights is that each element of the Grassmann necklace can be viewed as a function from the propagators of the diagram to the vertices in the Grassmann necklace element.  This is captured in Definition~\ref{def I_i as a function}, and used throughout the paper.
